Experience Report: Haskell in Computational Biology

### 2012-03-08 mrfy_experience_report.tex
Some further edits to AG's writing

As inexperienced functional programmers, developing a complex piece of
software in Haskell posed several barriers.
\remark{NOAH: You did the profiling, could you make sure what I've said here
is accurate and/or relevant? I didn't know what else to say. ---AG}
The first of which was profiling with GHC.
We previously mentioned that profiling with GHC produced ugly code with
cost center annotations scattered throughout our code, but we also found it
very difficult to profile code at all.
In particular, profiling required that all installed libraries be re-compiled
with profiling support---which was a process that we could not replicate
consistently.

Another barrier we faced was how to test our code using QuickCheck.
Indeed, the barrier was so formidable in our case that we still have not been 
able to use it.
The problems we faced specifically were two-fold: How do we start testing
a complex module? and How can we use QuickCheck to generate random inputs with 
complex constraints?

@@

\subsection{What you must know to succeed}
\remark{I don't like this. I really struggled with this section. ---AG}
If others would like to follow in our footsteps, there are a few
things we believe one must know in order to succeed.
Firstly, some experience with functional programming concepts is a necessity. 
Specifically, coming from an imperative background, writing algorithms and 
modeling data without the ability to mutate state were the most difficult 
concepts to adapt to.

Secondly, identifying existing tools and libraries from Hackage that are mature 
is also important.
We found that some packages that could have been immensely useful---like one 
that provides backtraces---were either not maintained or contained bugs.
Some other packages like Pads were functional but immature, which resulted in 
additional code to work around the parts of Pads that had not been implemented 
yet.

@@

\subsection{Evaluation}
\remark{I think we need more in this section. ---AG}
In hindsight, we feel that there are several key facets of our project in 
Haskell that deserve evaluation.
Firstly, we found that existing tools and libraries were plentiful and 
typically helped us greatly.
In particular, Data.Memocombinators made dynamic 
programming with Viterbi exquisitely simple and elegant.
Control.Parallel.Strategies also made the parallelization of Viterbi trivial.
Most of the tools we used came straight from Hackage, and although 
documentation of some form was almost always present, it was sometimes 
difficult to tell if a particular tool was right for us.
In particular, parts of the documentation were inaccessible to us as beginning 
functional programmers.
Documentation that was accessible to us typically contained straight-forward 
examples showing how to use that particular tool or library.

>>>>

As inexperienced functional programmers developing a complex piece of
software in Haskell, we encountered some barriers.
The first of these was profiling with GHC.
We previously mentioned that profiling with GHC required annotating
cost centers by hand, producing ugly code.
Beyond that, we also found it very difficult to profile code at all.
In particular, profiling required that all installed libraries be re-compiled
with profiling support,which was a process that we could not replicate
consistently.

Another barrier we faced was how to test our code using QuickCheck.
Indeed, the barrier was so formidable in our case that we still have not been 
able to use it.
The problems we faced specifically were two-fold: How do we start testing
a complex module? How can we use QuickCheck to generate random inputs with 
complex constraints? 
We wished to test invariants around beta-strand placement 
(potential solutions in our stochastic search), but found the QuickCheck generator
to be difficult to write.

@@

\subsection{What you must know to succeed}
If others would like to follow in our footsteps, there are a few
things we believe one must know in order to succeed.
Firstly, some experience with functional programming concepts is a necessity. 
Specifically, coming from an imperative background, we found transforming input data 
from the HMMER file format without the ability to mutate state was difficult.
In contrast, writing ``pure math'' like the Viterbi algorithm felt easier
and more natural than it would have been in C++.

Secondly, identifying existing tools and libraries from Hackage that are mature 
is also important.
We found that some packages that could have been immensely useful---like one 
that provides backtraces---were either not maintained or contained bugs.
Some other packages like Pads were functional but immature, which resulted in 
additional code to work around the parts of Pads that had not been implemented 
yet. 
In contrast, other packages such as \texttt{Data.Memocombinators} were
useful and mature.

@@

\subsection{Evaluation}
\remark{I think we need more in this section. ---AG}
In hindsight, we feel that there are several key facets of our project in 
Haskell that deserve evaluation.
Firstly, we found that existing tools and libraries were plentiful and 
typically helped us greatly.
In particular, \texttt{Data.Memocombinators} made dynamic 
programming with Viterbi exquisitely simple and elegant.
\texttt{Control.Parallel.Strategies} also made the parallelization of Viterbi trivial.
Most of the tools we used came straight from Hackage, and although 
documentation of some form was almost always present, it was sometimes 
difficult to tell if a particular tool was right for us.
In particular, parts of the documentation were inaccessible to us as beginning 
functional programmers.
Documentation that was accessible to us typically contained straight-forward 
examples showing how to use that particular tool or library.
